## Identify signature genesets for N, ACT, and ACT IL21 samples
Xulong Wang (xulong.wang@jax.org)

***

```{r, echo=FALSE, warning=FALSE}
library(ape)
library(amap)
library(ggplot2)
library(lattice)
library(ggdendro)
library(grid)

rm(list = ls())
```

##### TPM estimation of 38216 genes was used to identify the signature genesets. 
```{r}
setwd("~/Dropbox/GitHub/Lupus")
load("./data/myTpm.rdt")
dim(myTpm)
```

##### Only the genes whose maximal expression bigger than 50 and ANOVA p-value smaller than 0.05 will be considered. The ANOVA test gave variable genes among the samples, but not between the replicates. We have 865 genes for the downstream analysis.
```{r}
expr <- myTpm[apply(myTpm, 1, function(x) max(x) > 50), ]
group <- gsub("[12]", "", colnames(expr))
pval <- apply(expr, 1, function (x) min(summary(aov(x ~ group))[[1]][["Pr(>F)"]], na.rm = T))
expr <- expr[pval < 0.05, ]
dim(expr)
```

##### Gene expression in TPM was transformed into the log2 scale. The histogram below shows how the data looks like.
```{r}
expr <- log2(expr + 1)
hist(c(as.matrix(expr)));
```

##### We defined signature genes in two steps: (1) Each gene was assigned to either of the 3 groups from its maximal expression. (2) We conducted pair-wise comparsion between each gene's designated group and either of the other 2 groups. The genes that passed both t-test with p-value below 0.05 will be called as a signature gene of its designated group.
```{r}
emean <- data.frame("NN"=rowMeans(expr[1:2]), "NP"=rowMeans(expr[3:4]), "PP"=rowMeans(expr[5:6]))
idx.sample = apply(emean, 1, which.max)

profile1 <- profile2 <- list()
group <- c("1", "1", "2", "2")
for (sample in c("NN", "NP", "PP")) {
  
  if (sample == "NN") {
    expr0 <- expr[idx.sample == 1, ]
    expr01 <- expr0[, 1:4]
    expr02 <- expr0[, c(1:2, 5:6)]
    expr03 <- expr0[, 3:6]
  } else if (sample == "NP") {
    expr0 <- expr[idx.sample == 2, ]
    expr01 <- expr0[, 1:4]
    expr03 <- expr0[, c(1:2, 5:6)]
    expr02 <- expr0[, 3:6]
  } else {
    expr0 <- expr[idx.sample == 3, ]
    expr03 <- expr0[, 1:4]
    expr01 <- expr0[, c(1:2, 5:6)]
    expr02 <- expr0[, 3:6]
  }
  
  lm1 <- apply(expr01, 1, function(x) lm(x ~ group))
  lm2 <- apply(expr02, 1, function(x) lm(x ~ group))
  lm3 <- apply(expr03, 1, function(x) lm(x ~ group))
  
  idx1 <- lapply(lm1, function(x) summary(x)$coefficients[2, "Pr(>|t|)"] < 0.05)
  idx2 <- lapply(lm2, function(x) summary(x)$coefficients[2, "Pr(>|t|)"] < 0.05)
  idx3 <- lapply(lm3, function(x) summary(x)$coefficients[2, "Pr(>|t|)"] > 0.05)
  
  idx.gene1 <- unlist(idx1) & unlist(idx2)
  idx.gene2 <- unlist(idx1) & unlist(idx2) & unlist(idx3)
  profile1[[sample]] = rownames(expr0)[idx.gene1]
  profile2[[sample]] = rownames(expr0)[idx.gene2]
}
```

##### SCATTERPLOT ON PROFILIC GENES ---
```{r}
geneId <- unlist(profile1)
scatter <- myTpm[apply(myTpm, 1, function(x) max(x) > 5), ]
scatter <- log2(scatter + 1)
scatter <- data.frame("NN" = rowMeans(scatter[1:2]), "NP" = rowMeans(scatter[3:4]), "PP" = rowMeans(scatter[5:6]))

scatter$NP.NN <- scatter$NP - scatter$NN
scatter$PP.NN <- scatter$PP - scatter$NN
scatter$group <- rep("None", nrow(scatter))
scatter$group[rownames(scatter) %in% profile1[[1]]] <- "N"
scatter$group[rownames(scatter) %in% profile1[[2]]] <- "ACT"
scatter$group[rownames(scatter) %in% profile1[[3]]] <- "ACT IL21"

scatter1 <- scatter[scatter$group == "None", ]
scatter2 <- scatter[! scatter$group == "None", ]
scatter2$group <- factor(scatter2$group, levels = c("N", "ACT", "ACT IL21"))

hlId <- c("Foxo1", "Il4ra", "Sell", "Bcl6", "Btla", "Cd28", "Cxcr5", "Il21", "Maf")
myLabel <- scatter2[hlId, ]

ggplot() + 
  geom_point(data = scatter1, aes(x = NP.NN, y = PP.NN), color = "grey70", shape = 4, size = 1.5) +
  geom_point(data = scatter2, aes(x = NP.NN, y = PP.NN, color = group, shape = group), size = 1.5) +
  theme_bw() + xlab("ACT vs N") + ylab("ACT IL21 vs N") +
  scale_color_manual(values = c("darkorchid2", "chartreuse3", "dodgerblue3")) +
  theme(panel.border = element_blank(),
        axis.line = element_line(color = "grey30"),
        axis.text = element_text(size = 10),
        legend.title = element_blank(), 
        legend.text = element_text(size = 12),
        legend.key = element_blank()) +
  geom_point(data = myLabel, aes(x = NP.NN, y = PP.NN), shape = 2, size = 3, color = "red") 
```

##### PHYLO GRAPH ON SAMPLE HC 
```{r}
dt <- expr[geneId, ]
colnames(dt) <- c("N S1", "N S2", "ACT S1", "ACT S2", "ACT IL21 S1", "ACT IL21 S2")
hc1 <- hcluster(t(dt), method = "pearson", link = "average")
hc2 <- hcluster(dt, method = "correlation", link = "centroid")
hc1$labels <- gsub(" S[12]", "", hc1$labels)

par(mar = c(5, 2, 4, 2))
col.manual <- c("darkorchid2", "chartreuse3", "dodgerblue3", "grey20")
clusts = cutree(hc1, 3)
plot(as.phylo(hc1), tip.color = col.manual[clusts], 
     edge.color = col.manual[c(4, 1, 1, 4, 4, 2, 2, 4, 3, 3)],
     edge.width = 3, srt = 90, adj = 0.8, cex = 1, font = 2, 
     direction = "downward", label.offset = -5e-4)
```

##### HEATMAP ON PROLIFIC GENES ---
```{r}
tile1 <- dt[hc2$order, hc1$order]
tile2 <- t(apply(tile1, 1, scale))
colnames(tile2) <- colnames(tile1)
tile2 <- data.frame(value = c(tile2), gene = rep(rownames(tile1), 6), 
                    sample = rep(colnames(tile1), each = nrow(tile1)))
tile2$sample <- factor(tile2$sample, levels = colnames(tile1))
tile2$gene <- factor(tile2$gene, levels = rownames(tile1))

ggplot(tile2, aes(x = sample, y = gene, fill = value)) + 
  geom_tile() + guides(alpha = F) + coord_flip() +
  scale_fill_gradient2(low = "blue", high = "red") +
  theme_bw() + xlab("") + ylab("") + 
  theme(axis.text.x = element_text(size = 2, angle = 90),
        axis.text.y = element_text(size = 10),
        legend.text = element_text(size = 10), 
        legend.title = element_blank())
```